---
title: "Untitled"
output: html_document
date: "2026-02-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load packages
pacman::p_load(tidyverse)
```


We set up two types of agents. One is biased agent. It plays $1$ with fixed probability $p$. It is just a bernoulli draw. 

```{r}
biased_agent <- function(p){
  
  #return bernoulli flip
  choice <- rbinom(1,1,p)
  
  return(choice)
}
```


The second agent is a determinstic reinforcement learning agent. It keeps track of expected value of each choice on each trial. Since the there are only two choices with symmetric payoffs, our implementation only tracks the EV of one action. 
```{r}
# where v is expected value of playing right
rl_agent_no_soft_max <- function(v_prev, choice_prev, outcome_prev, lr) {

  ## Get the right PE ## 
  #if last choice was right just the outcome as feedback
  if (choice_prev == 1){
    feedback = outcome_prev
  } 
  
  #otherwise flip outcome, so its used 
  else {
    feedback = 1 - outcome_prev
  }
  
  #get pe
  PE = feedback - v_prev
  
  #update expected value of playing right
  v_curr <- v_prev + lr*PE
  
  
  choice <- rbinom(1, 1, v_curr)
  
  
  #return results as named values
  return(list(choice = choice, v_curr = v_curr))
}
```


```{r}
#softmaxxxing
softmax <- function(x, inv_temp){
  
  #get denominator
  denominator_sum = sum(exp(x*inv_temp))
  
  #for each val get softmax value and output
  output_vec = c()
  
  #loop over vars of expected vals
  for (i in 1:length(x)){
    prob_i <- (exp(x[i] * inv_temp))  / denominator_sum
    
    #store outputs
    output_vec <- c(output_vec, prob_i)
  }
  return(output_vec)
}
  
rl_agent_softmax <- function(v_prev, choice_prev, outcome_prev, lr, inv_temp) {

  ## Get the right PE ## 
  #if last choice was right just the outcome as feedback
  if (choice_prev == 1){
    feedback = outcome_prev
  } 
  
  #otherwise flip outcome, so its used 
  else {
    feedback = 1 - outcome_prev
  }
  
  #get pe
  PE = feedback - v_prev
  
  #update expected value of playing right
  v_curr <- v_prev + lr*PE
  
  #apply softmax
  prop_right <- softmax(c(v_curr, v_curr))
  
  
  choice <- rbinom(1, 1, v_curr)
  
  
  #return results as named values
  return(list(choice = choice, v_curr = v_curr))
}

```


We then run 25 simulations. In each we pair a biased agent with a specific $p$ of choosing 1 against a RL agent with a specific learning rate, $\alpha$ . We do cross 5 $\alpha$s against 5 $p$s and have the agents play 100 trials of the penny matching game. 
```{r}
set.seed(1)

## Setup configs for simulation loop ## 
#test loop were RL agent is matcher, ie it gets a point for matching opponent play. 
trials = 100
sims = 25

biases = c(0.05, 0.1, .2, .4, .5)
learning_rates = c(.01, 0.05, 0.1, .25, .5)

#prep df
total_df <- tibble()

for (sim_i in sims){
  for (learning_rate in learning_rates){
    for (bias in biases){
      
    ## Prep vectors
    #prep vector with expected vals of playing 1 for the RL agent
    expected_values = c(1)
    
    #prep vector for matcher (RL agent)
    matcher_choices = c(1)
    
    #prep vector for mismatcher
    mismatcher_choices = c(biased_agent(bias))
    
    #prep outcomes vector for matcher
    outcomes_matcher = c(if_else(mismatcher_choices [1] == matcher_choices[1],  1, 0))
    
    #loop through trials
      for (t in 2:trials){
      
        ## Biased agent. MISMATCHER
        mismatcher_choice <- biased_agent(bias)
        mismatcher_choices <- c(mismatcher_choices, mismatcher_choice)
        
        ## RL part: MATCHER
        #get vals for previous trial for RL
        outcome_prev = outcomes_matcher[t-1]
        choice_prev = matcher_choices[t-1]
        expected_value_prev = expected_values[t-1]
        
        #make choice for rl agent
        result <- rl_agent_no_soft_max(expected_value_prev, choice_prev, outcome_prev, learning_rate)
        
        matcher_choice <- result$choice
        v_curr  <- result$v_curr
        
        #store choices and expected values
        matcher_choices <- c(matcher_choices, matcher_choice)
        expected_values <- c(expected_values, v_curr)
      
        #get outcome at current trial and store
        outcome_matcher <- if_else(matcher_choice == mismatcher_choice, 1, 0)
        outcomes_matcher <- c(outcomes_matcher, outcome_matcher)
      } 
    
    #store a signle run as df
    df <- tibble(trial = c(1:trials), sim = rep(sim_i, trials), matcher_choices, mismatcher_choices, outcomes_matcher, expected_values, learning_rate = rep(learning_rate, trials), bias = rep(bias, trials)) 
    
    
    total_df <- bind_rows(total_df, df)
    }
  }
}
```

```{r}

learning_rate = 0.2

    ## Prep vectors
    #prep vector with expected vals of playing 1 for the RL agent
    expected_values = c(.5)
    
    #prep vector for matcher (RL agent)
    matcher_choices = c(1)
    
    #prep vector for mismatcher
    mismatcher_choices = c(biased_agent(bias))
    
    #prep outcomes vector for matcher
    outcomes_matcher = c(if_else(mismatcher_choices [1] == matcher_choices[1],  1, 0))
    
    #loop through trials
      for (t in 2:trials){
      
        ## Biased agent. MISMATCHER
        mismatcher_choice <- biased_agent(bias)
        mismatcher_choices <- c(mismatcher_choices, mismatcher_choice)
        
        ## RL part: MATCHER
        #get vals for previous trial for RL
        outcome_prev = outcomes_matcher[t-1]
        choice_prev = matcher_choices[t-1]
        expected_value_prev = expected_values[t-1]
        
        #make choice for rl agent
        result <- rl_agent_no_soft_max(expected_value_prev, choice_prev, outcome_prev, learning_rate)
        
        matcher_choice <- result$choice
        v_curr  <- result$v_curr
        
        #store choices and expected values
        matcher_choices <- c(matcher_choices, matcher_choice)
        expected_values <- c(expected_values, v_curr)
      
        #get outcome at current trial and store
        outcome_matcher <- if_else(matcher_choice == mismatcher_choice, 1, 0)
        outcomes_matcher <- c(outcomes_matcher, outcome_matcher)
      } 
    
    #store a signle run as df
stan_df <- tibble(trial = c(1:trials), matcher_choices, mismatcher_choices, outcomes_matcher, expected_values, learning_rate = rep(learning_rate, trials), bias = rep(bias, trials)) 

```




This provides us with a dataframe, which we can use for plotting. We get the cumultative performance. for the RL agent
```{r}
total_df  <- total_df %>% 
  group_by(learning_rate, bias, sim) %>% 
  #get the performance at each trial
  mutate(cumultative_performance = cumsum(outcomes_matcher))


#average
avg_total_df <-  total_df %>% 
  group_by(learning_rate, bias, trial) %>% 
  summarize(average_ev = mean(expected_values),
            average_cumulatative_performance_per_trial = mean(cumultative_performance / trial))

``` 


```{r}
#plot expected values

ggplot(avg_total_df, aes(x = trial, y=average_ev, color = bias, group = bias)) + 
  geom_line() +
  facet_wrap(~learning_rate) + 
  labs(y = "Expected Value", x = "Trial", ) +
  theme_classic()


```

```{r}
ggplot(avg_total_df, aes(x = trial, y=average_cumulatative_performance_per_trial , color = bias, group = bias)) + 
  geom_line() +
  facet_wrap(~learning_rate) +
  labs(y = "Performance", x = "Trial") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() 

```

