---
title: "Untitled"
output: html_document
date: "2026-02-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load packages
pacman::p_load(tidyverse)
```


We set up two types of agents. One is biased agent. It plays $1$ with fixed probability $p$. It is just a bernoulli draw. 

```{r}
biased_agent <- function(p){
  
  #return bernoulli flip
  choice <- rbinom(1,1,p)
  
  return(choice)
}

```


The second agent is a determinstic reinforcement learning agent. It keeps track of expected value of each choice on each trial. Since the there are only two choices with symmetric payoffs, our implementation only tracks the EV of one action. 
```{r}


# where v is expected value of playing right
rl_agent <- function(v_prev, choice_prev, outcome_prev, lr) {

  ## Get the right PE ## 
  #if last choice was right, update PE directly
  if (choice_prev == 1){
    PE <- outcome_prev - v_prev
  } 
  
  #otherwise flip outcome so we calculate PE as if we had played right
  else {
    PE <- -1*outcome_prev - v_prev
  }
  
  #update expected value of playing right
  v_curr <- v_prev + lr*PE
  
  #if ev playing right is positive, return 1, else return 0.
  if (v_curr > 0) {
    choice <- 1
  }
  else {
    choice <- 0
  }
  
  #return results as named values
  return(list(choice = choice, v_curr = v_curr))
}
```

We then run 25 simulations. In each we pair a biased agent with a specific $p$ of choosing 1 against a RL agent with a specific learning rate, $\alpha$ . We do cross 5 $\alpha$s against 5 $p$s and have the agents play 100 trials of the penny matching game. 
```{r}

## Setup configs for simulation loop ## 
#test loop were RL agent is matcher, ie it gets a point for matching opponent play. 
trials = 100

biases = c(0.05, 0.1, .2, .4, .8)
learning_rates = c(.01, 0.05, 0.1, .25, .5)



#prep df
total_df <- tibble()


for (learning_rate in learning_rates){
  for (bias in biases){
    
  ## Prep vectors
  #prep vector with expected vals of playing 1 for the RL agent
  expected_values = c(1)
  
  #prep vector for matcher (RL agent)
  matcher_choices = c(1)
  
  #prep vector for mismatcher
  mismatcher_choices = c(1)
  
  #prep outcomes vector for matcher
  outcomes_matcher = c(if_else(mismatcher_choices [1] == matcher_choices[1],  1, -1))
  
  
  #loop through trials
    for (t in 2:trials){
    
      ## Biased agent. MISMATCHER
      mismatcher_choice <- biased_agent(bias)
      mismatcher_choices <- c(mismatcher_choices, mismatcher_choice)
      
      ## RL part: MATCHER
      #get vals for previous trial for RL
      outcome_prev = outcomes_matcher[t-1]
      choice_prev = matcher_choices[t-1]
      expected_value_prev = expected_values[t-1]
      
      #make choice for rl agent
      result <- rl_agent(expected_value_prev, choice_prev, outcome_prev, learning_rate)
      
      matcher_choice <- result$choice
      v_curr  <- result$v_curr
      
      #store choices and expected values
      matcher_choices <- c(matcher_choices, matcher_choice)
      expected_values <- c(expected_values, v_curr)
    
  
      #get outcome at current trial and store
      outcome_matcher <- if_else(matcher_choice == mismatcher_choice, 1, -1)
      outcomes_matcher <- c(outcomes_matcher, outcome_matcher)
    } 
  
  #store a signle run as df
  df <- tibble(trial = c(1:trials), matcher_choices, mismatcher_choices, outcomes_matcher, expected_values, learning_rate = rep(learning_rate, trials), bias = rep(bias, trials)) 
  
  
  total_df <- bind_rows(total_df, df)
  }
}
```

This provides us with a dataframe, which we can use for plotting. 
```{r}
total_df
```
```{r}

```


